A Collection of Thoughts on the Keccak/SHA-3 Construction

### 2018-05-07 s2018-math490/thesis/thesis.tex
Last changes
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>
Updates to pdf

\begin{abstract}
    % TODO
\end{abstract}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Bergman and utilized
the resulting
\href{https://github.com/cipherboy/hash\_framework}{hash\_framework} project.
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.
A permanent location for this document is in the
\href{https://github.com/cipherboy/papers}{papers} repository.

@@

    Within cryptography's collection of algorithms, few are as useful as
hash functions have proven to be to cryptographers and non-cryptographers
alike. A hash function maps arbitrary length binary strings to binary strings
of a fixed length. Because the inputs are of arbitrary size, by the pigeonhole
principle, there must exist at least one collision. However, to be considered
cryptographically secure, finding such a collision must be hard (on the order
of $2^{\frac{b}{2}}$ for $b$ the number of bits in the output).

    Further, a cryptographic hash function must be resistant finding a
preimage: for a given output value $v$, finding an input $x$ such that
$f(x) = v$ should be roughly on the order of $2^{b}$. Given such an attack,
it is possible to find a collision: choose a random input, compute its hash,
then use the preimage attack to find a different input for that hash; these two
inputs form a collision pair. Thus a preimage attack is stronger than a
collision attack.

@@

@@

    $\theta$ is an eleven way XOR and is thus a linear map. For all
$0 \leq x \leq 4$ and $0 \leq z < w$, define:
\begin{align*}
    C[x, z] & := \oplus_{0 \leq y \leq 4} A[x, y, z] \\
    D[x, z] & := C[(x-1) \text{ mod } 5, z] \oplus C[(x+1) \text{ mod } 5, (z - 1) \text{ mod } w] \\
\end{align*}
Then:
\begin{align*}
    A'[x, y, z] & := A[x, y, z] \oplus D[x, z]
\end{align*}

    $\rho$ is a pure permutation function, permuting the locations of bits in
a lane (the $z$ component) of the state cube. For all
$0 \leq z < w: A'[0, 0, z] := A[0, 0, z]$. Then let $x, y = (0, 0)$, and for
$0 \leq t \leq 23$:
\begin{align*}
    \forall 0 \leq z < w: A'[x, y, z] & := A[x, y, (z - \frac{(t+1)(t+2)}{2}) \text{ mod } w] \\
    let x, y & = (y, (2x + 3y) \text{ mod } 5)
\end{align*}

@@

    $\chi$ is the only non-linear function and is of degree one (with inverse
of degree three). For all $0 \leq x \leq 4, 0 \leq y \leq 4, 0 \leq z < w$:
\begin{align*}
    A'[x, y, z] := A[x, y, z] \oplus (\lnot A[(x + 1) \text{ mod } 5, y, z] \land A[(x + 2) \text{ mod } 5, y, z]).
\end{align*}

@@

\subsection{Introduction to the Remaining Sections}

@@

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \oplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as strings of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$; thus
these are strings where bit $j = z' + 5\times x'$ is 1 if and only if
$C[x', z']$ is used to construct $C'[x, z]$.

(That is, since $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$).

>>>>

\begin{abstract}
    In this paper, we discuss new techniques for the cryptanalysis of
    Keccak/SHA-3, though none of the resulting attacks scale to full versions
    of the hash function. We show an explicit construction for the inverses of
    the core round functions, the permutation orders of the round functions,
    and an attack which exploits low-order cycles. Further, we introduce new
    logical cryptanalysis techniques and show the existence of partial fixed
    points to produce 2-1 block collisions.
\end{abstract}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Davis and utilized
the resulting
\href{https://github.com/cipherboy/hash\_framework}{hash\_framework} project
(viewable under the \texttt{cipherboy} account on GitHub.com).
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.
A permanent location for this document is in the
\href{https://github.com/cipherboy/papers}{papers} repository.

@@

    Within cryptography's collection of algorithms, few are as useful as
hash functions have proven to be to cryptographers and non-cryptographers
alike. A hash function maps arbitrary length binary strings to binary strings
of a fixed length; typically we denote such a function $h$. Because the inputs
are of arbitrary size, by the pigeonhole principle, there must exist at least
one collision: two blocks $x \neq y$ such that $h(x) = h(y)$. However, to be
considered cryptographically secure, finding such a collision must be hard
(on the order of $2^{\frac{b}{2}}$ for $b$ the number of bits in the output).

    Further, a cryptographic hash function must be resistant finding a
preimage: for a given output value $v$, finding an input $x$ such that
$h(x) = v$ should be roughly on the order of $2^{b}$. Given such an attack,
it is possible to find a collision: choose a random input, compute its hash,
then use the preimage attack to find a different input for that hash; these two
inputs form a collision pair. Thus a preimage attack is stronger than a
collision attack.

@@

\subsection{Introduction on the Usages of Hash Functions}

    Studying the integrity of hash function is important because hash functions
are a fundamental building block of both cryptographic and non-cryptographic
systems. While a body of literature developed around attacking the previous
hash functions (of MD4, MD5, SHA-1 and SHA-2), many of these attacks do not
directly translate to Keccak/SHA-3 because its construction differs
substantially. Further, many of these attacks started only as generic attacks
with low utility, and only with substantial work, were they translated to
highly specialized attacks on specific use cases. Below is a discussion of
various places where cryptographic hash functions are used, and what types of
attacks they are most susceptible to.

    The most common use case for hash functions are in signatures. Often times
the cryptography underlying asymmetric signatures (where one party can sign a
message and any party can verify that the signature and message is authentic)
can only sign fixed length messages, often only in the thousands of bytes.
However, longer data frequently must be signed. Thus, most implementations of
signature schemes use a cryptographically secure hash function and sign the
digest instead of the actual message. This can be seen in PGP, TLS, and PKI.
Often, these implementations are most vulnerable to a second preimage attack,
as one message is already provided. However, when the attacker has control
over both messages, a collision attack will suffice.

    Another common use case for hash functions are in message integrity
verification. In this case, a hash of the message is sent along side of the
message to ensure that the message hasn't been tampered with. While often this
signature should be signed, it is frequently used in cases where the
authenticity of the hash isn't questioned. This includes verifying contents of
files on disk or in version control systems such as Git. In both of these
cases, second preimage and collision attacks are the most useful for an
attacker.

\subsection{Introduction on the Scope of the Independent Study}

@@

    $\theta$ is an eleven way XOR and is thus a linear map. For all
$0 \leq x \leq 4$ and $0 \leq z < w$, define:
\begin{align*}
    C[x, z] & := \bigoplus_{0 \leq y \leq 4} A[x, y, z] \\
    D[x, z] & := C[(x-1) \text{ mod } 5, z] \oplus C[(x+1) \text{ mod } 5, (z - 1) \text{ mod } w] \\
\end{align*}
Then:
\begin{align*}
    A'[x, y, z] & := A[x, y, z] \oplus D[x, z]
\end{align*}

    $\rho$ is a pure permutation function, permuting the locations of bits in
a lane (the $z$ component) of the state cube. For all
$0 \leq z < w: A'[0, 0, z] := A[0, 0, z]$. Then let $(x, y) = (1, 0)$, and for
$0 \leq t \leq 23$:
\begin{align*}
    \forall 0 \leq z < w: A'[x, y, z] & := A[x, y, (z - \frac{(t+1)(t+2)}{2}) \text{ mod } w] \\
    \text{ let } (x, y) & = (y, (2x + 3y) \text{ mod } 5)
\end{align*}

@@

    $\chi$ is the only non-linear function and is of degree two (with inverse
of degree three). For all $0 \leq x \leq 4, 0 \leq y \leq 4, 0 \leq z < w$:
\begin{align*}
    A'[x, y, z] := A[x, y, z] \oplus (\lnot A[(x + 1) \text{ mod } 5, y, z] \land A[(x + 2) \text{ mod } 5, y, z]).
\end{align*}

@@

%\subsection{Introduction to the Remaining Sections}

@@

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \bigoplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \bigoplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as strings of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$; thus
these are strings where bit $j = z' + 5\times x'$ is 1 if and only if
$C[x', z']$ is used to construct $C'[x, z]$.

(That is, since $C'[x, z] = \bigoplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \bigoplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \bigoplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$).
